Attrition in Massive Open Online Courses: Inequalities, Dropout Reasons, and Psychological Factors

### 2014-10-30 paper_v1.tex
started general discussion, need to eat
final additions and edits before sumission

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different world: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}. Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 20,879 online learners in 18 courses. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

\begin{table*}[th]
\label{tab:s1sum}
\caption{Summary statistics of learner demographics and behavior in 14 MOOCs}
\small
\center
\begin{tabular}{llccccccccccc}
% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\toprule
 &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
 \cmidrule(r){4-6} \cmidrule(r){11-13}
 & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
\midrule
C1 & Business & 61,096   & 9,943  & 2,367    & N    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
C2 & Business & 41,081   & 6,544  & 883      & N    & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
C3 & CS       & 27,104   & 6,104 & 2,854     & N    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
C4 & CS       & 56,154   & 3,406 & 1,123     & N    & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
C5 & CS       & 57,636   & 5738  & 402       & N    & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
C6 & Engineering & 16,220 & 2,540 & 269      & N    & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
C7 & Enginerring & 7,178 & 1,312 & 202       & N    & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
C8 & Humanities & 7,634 & 3,631  & 152       & N    & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
C9 & Humanities & 6,475 & 3,315  & 326       & N    & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
C10 & Humanities & 9,903 & 4,227 & 1,570     & N    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
C11 & Humanities & 630 & 167     & 80        & N    & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
C12 & Math \& Stats & 49,317 & 3,090 & 862   & N    & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & N   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
C14 & Math \& Stats & 16,857 & 2,892 & 363   & N    & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & N   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
C16 & Math \& Stats & 27,236 & 2,711 & 637   & N    & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
C17 & Math \& Stats & 13,132 & 3,629 & 843   & N    & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
\bottomrule
\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelors' or higher degree  $^3$ self-reported in the feedback survey}
\end{tabular}
\end{table*}

@@

\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model for predicting persistence and achievement milestones (prop]{Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model for predicting persistence and achievement milestones (prop. videos watched, prop. assessments completed, and grade percentile). The baseline (intercept) are males in N. America of average age who have not earned a bachelors', no prior experience in the topic, not intending to complete the course, prepared to spend under 2 hours/week, this being their first MOOC. An estimate of -20 indicates a 20\% points lower probability of achieving a milestone.\label{fig:s1coefs}}
\end{figure*}

@@

\subsubsection{Satisfaction: Personal Success}
% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?

@@

The objective was to build an `early warning' prediction model that could flag learners who are likely to drop out in the future. We defined the following three criteria by which to evaluate the performance of the model.

\textit{Prediction accuracy}: Prediction recall (the fraction of correctly predicted dropouts) should be high to deliver the survey to as many future dropouts as possible. The false positive rate (the fraction of non-dropouts incorrectly predicted as droputs) should be low to avoid sending a survey to learners who do not experience challenges in the courses.

\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as being likely to drop out should be short. A shorter prediction lag improves the likelihood of successfully intervening to help the learner get back on track.

\textit{Transferability}: Whether or not a learner drops out of the course can not be determined before the end of the course. We thus had to train the prediction model using data from previous courses. The performance of the prediction model should be adequate in courses other than those on which it was trained.

The training set was composed of learners' interaction data (predictors) and dropout states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$). A second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model). The model was fit by logistic regression. Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).

\begin{table}[h!]
\caption{Performance of the dropout prediction model}
\label{tab:drop_pred_rocs}
\small
\center
\begin{tabular}{lccc}
\toprule
 & $train$ & $test_1$ & $test_2$ \\
\midrule
AUC &  0.931 & 0.929 & 0.922  \\
\bottomrule
\end{tabular}
\end{table}

Two of the predictors in the fitted model are highly influential. The first is the number of days since the learner has last been active ($x_{1}$). The likelihood that the learner drops out increases substantially if the learner disengages for 14 days or more. However, the likelihood of such a learner re-engaging increases with the fraction of released videos the learner has viewed ($x_{2}$) prior to the absence; this is the second most influential feature in the model. Equation 1 is an approximation of the model with only these two predictors. A learner was flagged if their dropout probability exceeded a half:

@@

Out of the 20,048 enrolled learners in the course, we only considered those 10,510 who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out. The survey was sent out to 6,053 leaners (excluding 21 learners with invalid email addresses or duplicate accounts).

@@

Study 2 focused on reasons for attrition and psychological factors of learners who were likely to drop out of the course. Through iterative coding, we developed a set of challenges that were faced by a subset of learners in a particular MOOC (cf. Table \ref{tab:s2reas}). The large majority of learners who indicate that they had not enough time was found to be separable into two groups: those with higher priority obligations that take up too much time, and those who are challenged by low levels of volitional control ($H4a$). MOOCs are an environment that requires learners to be highly self-regulated to be successful. The present findings emphasize the critical role of their capacity for volitional control, a factor that warrants further investigation \cite{corno2001volitional}. Very few learners explicitly indicated issues with volitional control in their responses, potentially because it is not socially desirable to admit. The absence of specific reasons for their shortage of time was found indicative of issues with volitional control ($H4b$). This could be a useful proxy for volitional issues in future research.

@@

\section{Acknowledgments}
% We thank Elise Ogle and Ruth Bram for helping with the development of the codebook. 
% Thank Emily for feedback and Andreas for feedback and data support.
Omitted for blind review.

>>>>

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}. Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses overall. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

\begin{table*}[th]
\label{tab:s1sum}
\caption{Summary statistics of learner demographics and behavior in 17 MOOCs}
\small
\center
\begin{tabular}{llccccccccccc}
% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\toprule
 &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
 \cmidrule(r){4-6} \cmidrule(r){11-13}
 & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
\midrule
C1 & Business & 61,096   & 9,943  & 2,367    & 1,282    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
C2 & Business & 41,081   & 6,544  & 883      & 630      & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
C3 & CS       & 27,104   & 6,104 & 2,854     & 2,116    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
C4 & CS       & 56,154   & 3,406 & 1,123     & 352      & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
C5 & CS       & 57,636   & 5738  & 402       & 342      & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
C6 & Engineering & 16,220 & 2,540 & 269      & 182      & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
C7 & Enginerring & 7,178 & 1,312 & 202       & 167      & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
C8 & Humanities & 7,634 & 3,631  & 152       & 127      & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
C9 & Humanities & 6,475 & 3,315  & 326       & 262      & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
C10 & Humanities & 9,903 & 4,227 & 1,570     & 1,066    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
C11 & Humanities & 630 & 167     & 80        & 40       & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
C12 & Math \& Stats & 49,317 & 3,090 & 862   & 510      & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 4,096   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
C14 & Math \& Stats & 16,857 & 2,892 & 363   & 282      & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 1,521   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
C16 & Math \& Stats & 27,236 & 2,711 & 637   & 610      & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
C17 & Math \& Stats & 13,132 & 3,629 & 843   & 626      & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
\bottomrule
\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelor's or higher degree  $^3$ self-reported in the feedback survey}
\end{tabular}
\end{table*}

@@

\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop]{Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop. videos watched, prop. assessments completed, and grade percentile). The baseline group were male learners in North America of average age who had not earned a bachelor's degree, without prior experience in the topic or other MOOCs, not intending to complete the course, and prepared to spend under two hours per week on it. An estimate of -20 indicates a 20\% points lower probability of achieving a milestone.\label{fig:s1coefs}}
\end{figure*}

@@

\subsubsection{Learner Satisfaction}
% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?

@@

We constructed an `early warning' prediction model to flag learners who had disengaged from the course or were about to disengage. The following three criteria were chosen as benchmarks to evaluate the practical and theoretical value of our prediction model.%   In order for our predictions to be useful, we need to observe the following three features of the dropout prediction problem.

\textit{Prediction accuracy}: Prediction recall (fraction of correctly predicted disengaged learners) should be high to deliver the survey to as many disengaged learners as possible. The false positive rate (fraction of active learners who are incorrectly predicted as disengaged) should be low to avoid sending surveys to learners who do not experience challenges in the course.

\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as disengaged should be short. A shorter prediction lag improves the likelihood of successfully intervening to help learners resume the course.

\textit{Transferability}: Whether or not a learner disengages from the course cannot be determined before the end of the course. We thus trained the prediction model using data from previous courses. The performance of the prediction model should be good in courses other than those on which it was trained.

The training set was composed of learners' interaction data (predictors) and disengagement states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$). A second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see omitted for blind review%\cite{halawa2014dropout}
, for additional information on the prediction model).

To control the length of the prediction lag, the model was only provided with learners' activity data until they were absent for 14 consecutive days. In the training phase, if the learner disengaged but the model did not detect it within 14 days of their last activity, the case was counted as a false negative. This construction results in a model with features that all fall within the defined prediction lag period. The model was fit by logistic regression with two-fold learner-level and course-level cross-validation. Table \ref{tab:predAuc} provides information on the prediction performance, measured by the area under the ROC curve (AUC).

\begin{table}[h!]
\caption{Performance of the attrition prediction model}
\label{tab:predAuc}
\small
\center
\begin{tabular}{lccc}
\toprule
 & $train$ & $test_1$ & $test_2$ \\
\midrule
AUC &  0.931 & 0.929 & 0.922  \\
\bottomrule
\end{tabular}
\end{table}

Two of the predictors in the fitted model are highly influential. The first is the number of days since the learner has last been active ($x_{1}$). The likelihood that the learner drops out increases substantially if the learner disengages for 14 days or more. However, the likelihood of such a learner re-engaging increases with the fraction of released videos the learner has viewed ($x_{2}$) prior to the absence; this is the second most influential feature in the model. Equation 1 is an approximation of the model with only these two predictors. The fact that almost identical results were obtained for $test_1$ and $test_2$ suggest good transferability of the model. A learner was flagged if their dropout probability exceeded a half:

@@

Out of the 20,048 enrolled learners in the course, only those 10,510 were considered who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out. The survey was sent out to 6,053 learners (excluding 21 learners with invalid email addresses or duplicate accounts).

@@

Study 2 focused on reasons for attrition and psychological factors of learners who were likely to drop out of the course. Through iterative coding, we developed a set of challenges that were faced by a subset of learners in a particular MOOC (cf. Table \ref{tab:s2reas}). The large majority of learners who indicate that they had not enough time was found to be separable into two groups: those with higher priority obligations that take up too much time, and those who are challenged by low levels of volitional control ($H4a$). MOOCs are an environment that requires learners to be highly self-regulated to be successful. The present findings emphasize the critical role of their capacity for volitional control, a factor that warrants further investigation \cite{corno2001volitional}. Very few learners explicitly indicated issues with volitional control in their open responses, potentially because it is not socially desirable to admit. The absence of specific reasons for their shortage of time was found indicative of issues with volitional control ($H4b$). This could be a useful proxy for volitional issues in future research.

@@

The present studies extend the empirical base of research on attrition in online learning environments. A large-scale study of over 60,000 online learners revealed strong individual differences in persistence and performance by demographic and geographic features. Learners who reported that they dropped out were less satisfied with their course achievements. Time was a critical obstacle for many learners, and a low capacity for volitional control was found to contribute to this issue, though learners seemingly avoid admitting this. Successful online learners, according to their own judgment, were found to exhibit higher levels of goal-striving, social belonging, and growth mindset than subjectively unsuccessful learners. We also began to develop a survey item for measuring reasons for disengaging from a course (cf. Table \ref{tab:s2reas}), which should be tested and extended, if needed, based on more empirical evidence from additional MOOCs.

This work yields a number of theoretical and practical implications. In addressing research questions and testing hypotheses which emerged from a review of the literature, we sought to advance theory on attrition and provide empirical evidence. In line with early models of attrition that emphasized the importance of goal-directed behavior, we found goal striving to be higher in subjectively successful than unsuccessful learners. Moreover, volitional control was identified as a major contributor in reported time-related challenges with taking a MOOC. Tinto's \cite{tinto1975dropout} classic model emphasizes the role of academic and social integration, and also builds on various learner characteristics as contributors to learners' decision to persist in a course of study. Our findings provide strong support for the critical role of individual characteristics, such as gender, age, intentions, experiences, and geographic location; and, leveraging the more recent notion of social belonging \cite{walton2007question}, we found evidence in support of Tinto's social integration component. Parallel to positive attributions of attrition in community college settings, we found progress in the MOOC and learner satisfaction to only be weakly related: learners who watched very few lectures were almost as satisfied as those who watched most lectures. Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers.

There are many practical implications of this research and the one we would like to emphasize is the need for and feasibility of targeted psychological interventions in online learning environments. While the current work does not provide explanations for the presence of substantial achievement gaps, we can speculate that similar psychological processes as in traditional educational environments are involved. Brief and targeted online interventions (e.g., mindset, social belonging, self-affirmation, goal striving) can potentially narrow existing achievement gaps in online learning. We hope to see more research on developing accurate prediction models that identify struggling learners and suitable interventions to best support them.

\section{Acknowledgments}
% We thank Elise Ogle and Ruth Bram for helping with the development of the initial codebook. 
% Thank Emily Schneider for her helpful feedback and Andreas P. for his feedback and data support.
Omitted for blind review.
